---
title: "CellCountModeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CellCountModeling}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width=7,
  fig.asp = 0.8,
  cache = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bestdr)
library(dplyr)
library(tidyr)
library(magrittr)
library(ggplot2)
library(cmdstanr)
library(estipop)
```

# Introduction
We introduce the basic workflow for running a dose response model for counting
total cell counts to get the birth and death rates. We will simulate data to
provide an example of how data should look for import and running in BESTDR. The
resulting model will show how BESTDR can be used to distinguish cytotoxic and
cytostatic effects. Second, we will simulate multiple "cell lines" to show how
hierarchical modeling can be used to model multiple cell line responses
simultaneously.


# Single Cell Line Estimation

## Data Simulation
We will generate parameters from a concentration-response curve for both the
birth and death rates. That is, cell will split with rate $b(c)$ and die with
rate $d(c)$. The net growth rate is the difference of these two, or
$g(c) = b(c) - d(c)$. The concentrations tested are at plotted as points in the
figure with the smooth curve representing the smooth function for the birth
(blue), death (red), and growth (purple) rates.

```{r data-generation}
# Generate curve from 4-parameter logistic function for birth and death
b0 <- 1 # left asymptote
bi <- 0.2 # right asymptote
b50 <- -5 # log-dose at midpoint of birthrate (should switch to log scale)
bh <- 2 # Hill Coefficient
db <- b0 - bi # Difference in asymptotes (for parameterization)

d0 <- 0.01
di <- 0.5
d50 <- 1
dh <- 2
dd <- di - d0

b_curve <- function(lc, bi, db, bh, b50) bi + db / (1 + exp(bh * (lc - b50)))
d_curve <- function(lc, d0, dd, dh, d50) d0 + dd - dd / (1 + exp(dh * (lc - d50)))

param_truth <- data.frame(
  bi = bi, b50 = b50, bh = bh, db = db,
  d0 = d0, d50 = d50, dh = dh, dd = dd
)

crange <- c(0, 10^seq(-4, 4, 0.1))
lc_range <- log(crange)
concentrations <- c(0, 10^seq(-4, 4, 1))
log_conc <- log(concentrations)

b_truth <- b_curve(lc_range, bi, db, bh, b50)
d_truth <- d_curve(lc_range, d0, dd, dh, d50)
g_truth <- b_truth - d_truth

curve_truth <- data.frame(concentration = crange,
                          log_conc = lc_range,
                          b = b_truth, d = d_truth, g = g_truth) %>%
  pivot_longer(-c(concentration, log_conc), names_to = "Rate", values_to = "Value")


curve_truth %>%
  ggplot(aes(x = concentration, y = Value, color = Rate)) +
  geom_line() +
  geom_point(data = curve_truth %>% filter(concentration %in% concentrations)) +
  scale_x_log10() +
  labs(x = "Concentration", y = "Rate") +
  scale_color_manual(values = c("darkblue", "darkred", "purple")) +
  theme_bw()
```

We simulate the process using the estipop package. The "rep" variable represents
the replicate and the "type1" variable represents the count at a given time.
```{r simulate}
N <- 10 # Number of samples
tot_time <- 3
dt <- 0.1 # Time between measurements
times <- seq(0, tot_time, dt)
init_pop <- 200 # Number of ancestors

rate_truth <- data.frame(concentration = crange,
                         log_conc = lc_range,
                         b = b_truth, d = d_truth, g = g_truth) %>%
  filter(concentration %in% concentrations)
# Simulate process with estipop
set.seed(9283)
dat <- data.frame()
for (i in seq_along(rate_truth$concentration)) {
  params <- c(rate_truth$b[i], rate_truth$d[i])
  # time-homogeneous two-type model
  model <- estipop::process_model(
    estipop::transition(rate = estipop::rate(params[1]), parent = 1, offspring = 2), # birth
    estipop::transition(rate = estipop::rate(params[2]), parent = 1, offspring = 0) # death
  ) # clearance
  
  dat_temp <- estipop::branch(model, params, init_pop, times, N, silent = T)
  dat_temp$concentration <- concentrations[i]
  dat_temp$log_conc <- log_conc[i]
  dat <- bind_rows(dat, dat_temp)
}

head(dat)

# Plot to look at data
dat %>% ggplot(aes(x = time, group = as.factor(rep))) +
  geom_line(aes(y = type1), col = "black") +
  facet_wrap(concentration ~ .) +
  theme_bw() +
  scale_y_log10()
```

The data needs to be preprocessed before using the model by taking the difference
between counts between successive time points.

```{r preprocess}
dat_dt <- dat %>%
  group_by(rep, concentration) %>%
  mutate(
    dt = signif(time - lag(time), 6),
    type1_prev = lag(type1)
) %>% filter(!is.na(dt))
```
## Estimation
We will use a Stan file from the model library to estimate the
birth and death rates. We will set up the necessary input data as a list and
concatenate the prior distribution parameters for the means and standard
deviations for each of the parameters to estimate.

The single type model takes the current and previous count as well as the
time between the two counts in separate vectors that are the length of the number
of observations (note: observations are not replicates, but the number of time points recorded
over all replicates). Each observation has a concentration, convert the concentrations
to an index that corresponds to a vector of unique concentrations. The concentration
index is used in place of the concentrations for each observation.

That is, if we have concentrations for a set of 5 observations (0.0, 0.1, 0.1, 0.25, 0.0),
then the `c_idx` would be (1, 2, 2, 3, 1), `log_conc` would be (0.0, 0.1, 0.25),
and `nc` would be 3.

```{r estimation-preprocess}
# Estimation with cmdstanr
stanfileloc <- system.file("model_library/1type-livecell/logistic/birthdeath.stan", package = "bestdr")
stanmodel <- cmdstan_model(stanfileloc)

# Set up data for inputs
dat_dt$log_conc_idx <- as.numeric(factor(dat_dt$log_conc, ordered = T))
log_conc <- unique(dat_dt$log_conc)
num_conc <- length(log_conc)

data.list <- list(
  N = nrow(dat_dt),
  count = dat_dt$type1,
  count_prev = dat_dt$type1_prev,
  dt = dat_dt$dt,
  c_idx = dat_dt$log_conc_idx,
  nc = num_conc,
  conc = log_conc
)
prior.list <- list(
  prior_mu_bi = 1,
  prior_s_bi = 0.3,
  prior_mu_db = 0.5,
  prior_s_db = 0.5,
  prior_mu_b50 = 0,
  prior_s_b50 = 1,
  prior_mu_bh = 1,
  prior_s_bh = 0.5,
  prior_mu_d0 = 0.3,
  prior_s_d0 = 0.3,
  prior_mu_dd = 0.5,
  prior_s_dd = 0.5,
  prior_mu_d50 = 0,
  prior_s_d50 = 1,
  prior_mu_dh = 1,
  prior_s_dh = 0.5
)
initfun <- function() {
  list(
    bi = runif(1, 0, 2),
    db = runif(1, 0, 2),
    b50 = runif(1, -2, 2),
    bh = runif(1, 0, 2),
    d0 = runif(1, 0, 2),
    dd = runif(1, 0, 2),
    d50 = runif(1, -2, 2),
    dh = runif(1, 0, 2),
    count_err = runif(data.list$N, -0.5, 0.5)
  )
}
```

For speed, we run the sampler for a single chain for 500 samples
```{r sampler, warning=FALSE}
model_posterior <- stanmodel$sample(
  seed = 117,
  data = c(data.list, prior.list),
  chains = 1,
  parallel_chains = 1,
  refresh = 0,
  init = initfun,
  iter_warmup = 1000,
  iter_sampling = 500
)
```

```{r posterior-summary, warning=FALSE}
model_posterior$summary()
params_posterior <- model_posterior$draws(variables = c("bi", "db", "b50", "bh", "d0", "dd", "d50", "dh"), format = "df")
```
## Analyzing Results

The following plot shows the posterior samples with the true parameter as a
vertical line.
```{r comparison, warning=FALSE}
# Compare to ground truth
param_truth_l <- param_truth %>% pivot_longer(everything(), names_to = "stat")

params_posterior_l <- params_posterior %>%
  select(.chain, bi, db, b50, bh, d0, dd, d50, dh) %>%
  pivot_longer(!c(".chain"), names_to = "stat")

# Marginal Posteriors
params_posterior_l %>%
  ggplot(aes(x = value, group = .chain)) +
  geom_density() +
  geom_vline(data = param_truth_l, aes(xintercept = value), linewidth = 2) +
  facet_wrap(. ~ stat, scales = "free") +
  theme_bw()
```
We additionally take posterior samples and create dose-response samples for each
to get confidence bands for the dose response curves.

```{r setup-posterior-curves}
n_posteriors <- 500
crange <- c(10^seq(-6, 4, 0.1))
lcrange <- log(crange)
sample_post <- sample(nrow(params_posterior), n_posteriors, replace = FALSE)
b_post <- data.frame(
  crange,
  sapply(sample_post, function(x) {
    b_curve(
      lcrange,
      params_posterior$bi[x],
      params_posterior$db[x],
      params_posterior$bh[x],
      params_posterior$b50[x]
    )
  })
) %>%
  pivot_longer(cols = !crange, names_to = "sample", values_to = "b")
d_post <- data.frame(
  crange,
  sapply(sample_post, function(x) {
    d_curve(
      lcrange,
      params_posterior$d0[x],
      params_posterior$dd[x],
      params_posterior$dh[x],
      params_posterior$d50[x]
    )
  })
) %>%
  pivot_longer(cols = !crange, names_to = "sample", values_to = "d")
posterior_samples <- b_post %>%
  inner_join(d_post) %>%
  mutate(crange = ifelse(crange == 0, 0.000001, crange),
         g = b - d)
head(posterior_samples)
```

```{r plotting-curves}
posterior_samples %>% pivot_longer(-c(crange, sample),
                                   names_to = "param", values_to = "value") %>%
  ggplot(aes(x = crange, y = value, color = param, group = interaction(param, sample))) +
  geom_line(alpha = 0.1, linewidth = 0.5) +
  scale_x_log10() +
  scale_color_manual(values = c("darkblue", "darkred", "purple")) +
  theme_bw()
```
The above curves show that the simulation we generated has initial cytostatic
effects noted by the decreasing birth rate followed by cytotoxic effects as the
concentration increases further which is based on the increasing death rate. We
were able to estimate each effect simulataneously.

## Additional metrics
We can further derive additional metrics about the birth, death, or growth curves.
For instance, the $b_{50}$ and $d_{50}$ are the concentration at which the birth
and death rates reach their midpoint which are parameters of the logistic curve
and can provide a measure of how cytotoxic or cytostatic an agent is.

For summary statistics, we can define the $NG_{50}$ and $NG_{0}$ as the
concentration where the growth rate hits half it's initial value and the
concentration when the growth rate hits 0 respectively. In the latter case,
this is when the cell growth starts decreasing over time.


# Multiple Cell Lines (Hierarchical Model)

## Data Simulation

Using data for a hierarchical model requires an additional group variable
to be included. We index this the same way we index concentrations. We'll
illustrate the process for estimating from multiple cell lines by sampling
parameters from distributions where we've specified the hyperparameters.

```{r hier-params, warning=FALSE}
# parameters
set.seed(1238)
b0 <- rnorm(5, 1, 0.05)
bi <- rnorm(5, 0.2, 0.02)
db <- b0 - bi
b50 <- rnorm(5, -5, 0.5)
bh <- rnorm(5, 2, 0.3)
d0 <- rnorm(5, 0.05, 0.02)
di <- rnorm(5, 0.5, 0.1)
dd <- di - d0
d50 <- rnorm(5, 0, 0.2)
dh <- rnorm(5, 2, 0.4)


param_truth <- data.frame(
  group = 1:length(b0),
  bi = bi, db = db, b50 = b50, bh = bh,
  d0 = d0, dd = dd, d50 = d50, dh = dh
)

concentrations <- c(0, 10^seq(-4, 4, 1))
log_conc <- log(concentrations)

b_curve <- function(lc, bi, db, bh, lb50) bi + db / (1 + exp(bh * (lc - lb50)))
d_curve <- function(lc, d0, dd, dh, ld50) d0 + dd - dd / (1 + exp(dh * (lc - ld50)))

b_truth <- data.frame(sapply(1:length(bi), function(i) b_curve(log_conc, bi[i], db[i], bh[i], b50[i])))
d_truth <- data.frame(sapply(1:length(d0), function(i) d_curve(log_conc, d0[i], dd[i], dh[i], d50[i])))
b_truth$concentration <- concentrations
d_truth$concentration <- concentrations
b_truth <- b_truth %>% pivot_longer(!concentration, names_to="group", values_to="b")
d_truth <- d_truth %>% pivot_longer(!concentration, names_to = "group", values_to="d")
curve_truth <- b_truth %>%
  inner_join(d_truth) %>%
  mutate(group = sub("X", "", group),
         g = b - d)

curve_truth %>%
  pivot_longer(-c(concentration, group), names_to = "param", values_to = "value") %>%
  ggplot(aes(x = concentration, y = value, color = param, group = interaction(param, group))) +
    geom_line() +
    geom_point(size = 2) +
  scale_color_manual(values = c("darkblue", "darkred", "purple")) +
  scale_x_log10() +
  theme_bw()
```

Just as before we simulate the data using estipop, but we also loop over the
groups / cell lines as well. In this case the `curve_truth` data frame already
has 1 row per concentration-group combination.

```{r hier-simulate}
N <- 5 # Number of samples
tot_time <- 3
dt <- 0.1 # Time between measurements
times <- seq(0, tot_time, dt)
init_pop <- 200 # Number of ancestors


# Simulate process with estipop
set.seed(342)
dat <- data.frame()
for (i in 1:nrow(curve_truth)) {
  params <- c(curve_truth$b[i], curve_truth$d[i])
  # time-homogeneous two-type model
  model <- estipop::process_model(
    estipop::transition(rate = estipop::rate(params[1]), parent = 1, offspring = 2), # birth
    estipop::transition(rate = estipop::rate(params[2]), parent = 1, offspring = 0) # death
  ) # clearance
  
  dat_temp <- estipop::branch(model, params, init_pop, times, N, silent = T)
  dat_temp$concentration <- curve_truth$concentration[i]
  dat_temp$log_conc <- log(dat_temp$concentration)
  dat_temp$group <- curve_truth$group[i]
  dat <- bind_rows(dat, dat_temp)
}
head(dat)

dat %>% ggplot(aes(x = time, group = interaction(group, as.factor(rep)))) +
  geom_line(aes(y = type1, color = group), alpha = 0.5) +
  facet_wrap(concentration ~ ., scales = "free_y") +
  scale_y_log10() +
  scale_color_manual(values = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#a6761d")) +
  theme_bw()
```

```{r hier-preproc}
# Preprocessing
dat_dr <- dat %>%
  group_by(rep, concentration, group) %>%
  mutate(
    dt = signif(time - lag(time), 6),
    type1_prev = lag(type1)
  ) %>%
  filter(!is.na(dt))

# Set up the data for input into Stan
dat_dr$log_conc_idx <- as.numeric(factor(dat_dr$log_conc, ordered = T))
log_conc <- unique(dat$log_conc)
num_conc <- length(log_conc)
num_groups <- length(unique(dat$group))
```

## Estimation
```{r hier-estimate, warning=FALSE, message=FALSE}
# Estimation with cmdstanr
stanfileloc <- system.file("model_library/1type-livecell/logistic/birthdeath_mixedeffects.stan",
                           package = "bestdr")
stanmodel <- cmdstan_model(stanfileloc)

data.list <- list(
  N = nrow(dat_dr),
  count = dat_dr$type1,
  count_prev = dat_dr$type1_prev,
  dt = dat_dr$dt,
  c_idx = dat_dr$log_conc_idx,
  nc = num_conc,
  conc = log_conc,
  g_idx = as.numeric(as.factor(dat_dr$group)),
  ng = num_groups
)
prior.list <- list(theta_1_mu_bi = 1.0,
                   theta_2_mu_bi = 0.4,
                   theta_1_s_bi = 0.5,
                   theta_2_s_bi = 0.2,
                   theta_1_mu_db = 1.0,
                   theta_2_mu_db = 0.4,
                   theta_1_s_db = 0.5,
                   theta_2_s_db = 0.2,
                   theta_1_mu_b50 = 0.0,
                   theta_2_mu_b50 = 0.5,
                   theta_1_s_b50 = 0.5,
                   theta_2_s_b50 = 0.2,
                   theta_1_mu_bh = 1.0,
                   theta_2_mu_bh = 0.5,
                   theta_1_s_bh = 0.5,
                   theta_2_s_bh = 0.2,
                   theta_1_mu_d0 = 0.1,
                   theta_2_mu_d0 = 0.4,
                   theta_1_s_d0 = 0.5,
                   theta_2_s_d0 = 0.2,
                   theta_1_mu_dd = 1.0,
                   theta_2_mu_dd = 0.4,
                   theta_1_s_dd = 0.5,
                   theta_2_s_dd = 0.2,
                   theta_1_mu_d50 = 0.0,
                   theta_2_mu_d50 = 0.5,
                   theta_1_s_d50 = 0.5,
                   theta_2_s_d50 = 0.2,
                   theta_1_mu_dh = 1.0,
                   theta_2_mu_dh = 0.5,
                   theta_1_s_dh = 0.5,
                   theta_2_s_dh = 0.2)
initfun = function() {
  list(
    bi = runif(num_groups, 0.1, 2),
    db = runif(num_groups, 0.1, 2),
    b50 = runif(num_groups, -2, 2),
    bh = runif(num_groups, 0.1, 2),
    d0 = runif(num_groups, 0.1, 2),
    dd = runif(num_groups, 0.1, 2),
    d50 = runif(num_groups, -2, 2),
    dh = runif(num_groups, 0.1, 2),
    count_err = runif(data.list$N, -0.5, 0.5)
  )
}

# Sampling
model_posterior <- stanmodel$sample(
  seed=711,
  data = c(data.list, prior.list),
  chains = 1,
  parallel_chains = 1,
  refresh = 0,
  init = initfun,
  iter_warmup = 1000,
  iter_sampling = 500
)
```

```{r hier-results}
params_posterior <- model_posterior$draws(
  variables =
    c("bi", "db", "b50", "bh", "d0", "dd", "d50", "dh"), format = "df"
)
hyperparams_posterior <- model_posterior$draws(
  variables =
    c("mu_bi", "mu_db", "mu_b50", "mu_bh", "mu_d0", "mu_dd", "mu_d50", "mu_dh",
      "s_bi", "s_db", "s_b50", "s_bh", "s_d0", "s_dd", "s_d50", "s_dh"), format = "df"
)

params_posterior_l <- params_posterior %>%
  select(.chain, .draw, starts_with("b"), starts_with("d")) %>%
  pivot_longer(!c(".chain", ".draw"), names_to = "stat") %>%
  separate(stat, sep = "\\[", c("stat", "group")) %>%
  mutate(group = (sub("\\]", "", group)))

params_posterior_l %>% group_by(stat, group) %>%
  summarize(value = mean(value)) %>%
  pivot_wider(names_from = "stat", values_from = "value")

ground_truth_l <- param_truth %>%
  pivot_longer(!group, names_to = "stat") %>%
  mutate(group = as.character(group))
params_posterior_l %>%
  ggplot(aes(x = value, color = group, group = interaction(group, .chain))) +
  geom_density() +
  geom_vline(data = ground_truth_l, aes(xintercept = value, color = group), linewidth = 2) +
  facet_wrap(stat ~ ., scales = "free") +
  theme_bw()
```

Finally, we plot the curves for concentration response posterior samples from
each of the 5 groups individually. 
```{r hier-posteriorcurves}
# Samples of posterior curves
n_posteriors <- 300
crange <- c(10^seq(-6, 4, 0.1))
lcrange <- log(crange)
posterior_samples <- data.frame()
for (i in 1:num_groups) {
  sample_post <- sample(nrow(params_posterior), n_posteriors)
  posterior_subset <- params_posterior_l %>%
    filter(group == i, .draw %in% sample_post) %>%
    select(-c(.chain, group)) %>%
    pivot_wider(id_cols = .draw, names_from = "stat")
  b_subset <- data.frame(
    group = i,
    crange,
    sapply(1:n_posteriors, function(x) {
      b_curve(
        lcrange,
        posterior_subset$bi[x],
        posterior_subset$db[x],
        posterior_subset$bh[x],
        posterior_subset$b50[x]
      )
    })
  ) %>%
    pivot_longer(cols = !c(group, crange), names_to = "sample", values_to = "b")
  d_subset <- data.frame(
    group = i,
    crange,
    sapply(1:n_posteriors, function(x) {
      d_curve(
        lcrange,
        posterior_subset$d0[x],
        posterior_subset$dd[x],
        posterior_subset$dh[x],
        posterior_subset$d50[x]
      )
    })
  ) %>%
    pivot_longer(cols = !c(group, crange), names_to = "sample", values_to = "d")
  posterior_samples <- posterior_samples %>%
    bind_rows(b_subset %>%
                inner_join(d_subset) %>%
                mutate(crange = ifelse(crange == 0, 0.000001, crange)))
}


posterior_samples %>%
  pivot_longer(-c(group, crange, sample), names_to = "param", values_to = "value") %>%
  ggplot(aes(x = crange, y = value, color = param, group = interaction(param, sample))) +
  geom_line() +
  facet_grid(.~group) +
  scale_color_manual(values = c("darkblue", "darkred")) +
  scale_x_log10() +
  theme_bw()
```